---
title: 'Can we get good estimates of the most abundant species by only counting some sections?'
output: word_document
---

```{r error=FALSE, message=FALSE, echo=FALSE, warning=FALSE}

library(tidyverse)
options(scipen = 999)


source("C:/Users/scott.jennings/Documents/Projects/water_birds/code/waterbird_utility_functions.r")
source("C:/Users/scott.jennings/Documents/Projects/R_general/utility_functions/bird_utility_functions.r")

# read cleaned data
wbirds4analysis <- readRDS("C:/Users/scott.jennings/Documents/Projects/water_birds/data_files/working_rds/wbirds4analysis") %>% 
  wbird_add_study_day() 


most_abund <- c("SUSC", "BUFF", "SCAUP", "GRSC", "RUDU", "HOGR", "BLBR", "AMCO", "DCCO", "WEGR", "BRAC", "AMWI" )


```



```{r error=FALSE, message=FALSE, echo=FALSE, warning=FALSE}

# could we get away with only surveying some spp in some sections? ----

wbird_baywide_spp_season_survey <- wbirds4analysis %>% 
  group_by(study.year, date, alpha.code, section) %>% 
  summarise(section.count = sum(section.final)) %>% 
  ungroup() %>% 
  group_by(study.year, date, alpha.code) %>% 
  mutate(baywide.count = sum(section.count)) %>% 
  arrange(alpha.code, date) %>% 
  ungroup()



```




Considering the species that were among the top 5 most abundant in at least 1 year, is there any section of the bay where the count of that section best represents the count of the entire bay?





For Scaup, section 2 appears most important.
```{r error=FALSE, message=FALSE, echo=FALSE, warning=FALSE}

wbird_baywide_spp_season_survey %>% 
  filter(alpha.code %in% c("GRSC", "SCAUP")) %>% 
  ggplot() +
  geom_point(aes(baywide.count, section.count, color = alpha.code)) +
  facet_wrap(~section) +
  ggtitle("GRSC, SCAUP") +
  theme_bw()

```

For Double-crested and Brandt's Cormorant, section 2 seems most important.

```{r error=FALSE, message=FALSE, echo=FALSE, warning=FALSE}

wbird_baywide_spp_season_survey %>% 
  filter(alpha.code %in% c("DCCO", "BRAC")) %>% 
  ggplot() +
  geom_point(aes(baywide.count, section.count, color = alpha.code)) +
  facet_wrap(~section)+
  ggtitle("DCCO, BRAC") +
  theme_bw()

```


For Bufflehead, section 3 appears most important.

```{r error=FALSE, message=FALSE, echo=FALSE, warning=FALSE}

wbird_baywide_spp_season_survey %>% 
  filter(alpha.code %in% c("BUFF")) %>% 
  ggplot() +
  geom_point(aes(baywide.count, section.count, color = alpha.code)) +
  facet_wrap(~section) +
  ggtitle("BUFF") +
  theme_bw()

```


For Surf Scoter, section 2 appears most important.

```{r error=FALSE, message=FALSE, echo=FALSE, warning=FALSE}

wbird_baywide_spp_season_survey %>% 
  filter(alpha.code %in% c("SUSC")) %>% 
  ggplot() +
  geom_point(aes(baywide.count, section.count, color = alpha.code)) +
  facet_wrap(~section) +
  ggtitle("SUSC") +
  theme_bw()

```

For Ruddy Duck, sections 1 and 3 appear most important, but there are some instances where the counts in these sections poorly match the baywide count.
```{r error=FALSE, message=FALSE, echo=FALSE, warning=FALSE}

wbird_baywide_spp_season_survey %>% 
  filter(alpha.code %in% c("RUDU")) %>% 
  ggplot() +
  geom_point(aes(baywide.count, section.count, color = alpha.code)) +
  facet_wrap(~section) +
  ggtitle("RUDU") +
  theme_bw()

```

For small grebes, section 2 appears most important.
```{r error=FALSE, message=FALSE, echo=FALSE, warning=FALSE}

wbird_baywide_spp_season_survey %>% 
  filter(alpha.code %in% c("HOGR", "EAGR", "HEGR")) %>% 
  ggplot() +
  geom_point(aes(baywide.count, section.count, color = alpha.code)) +
  facet_wrap(~section) +
  ggtitle("HOGR, EAGR") +
  theme_bw()

```


For Black Brant, sections 3 and 4 appear most important.Should count both sections for this spp.
```{r error=FALSE, message=FALSE, echo=FALSE, warning=FALSE}

wbird_baywide_spp_season_survey %>% 
  filter(alpha.code %in% c("BLBR")) %>% 
  ggplot() +
  geom_point(aes(baywide.count, section.count, color = alpha.code)) +
  facet_wrap(~section) +
  ggtitle("BLBR") +
  theme_bw()

```

For coots, sections 1 and 3 appear most important, but there are several instances where the counts in these sections did not match the baywide count.

```{r error=FALSE, message=FALSE, echo=FALSE, warning=FALSE}

wbird_baywide_spp_season_survey %>% 
  filter(alpha.code %in% c("AMCO")) %>% 
  ggplot() +
  geom_point(aes(baywide.count, section.count, color = alpha.code)) +
  facet_wrap(~section) +
  ggtitle("AMCO") +
  theme_bw()

```

For Western Grebe, section 2 appears most important.

```{r error=FALSE, message=FALSE, echo=FALSE, warning=FALSE}

wbird_baywide_spp_season_survey %>% 
  filter(alpha.code %in% c("WEGR")) %>% 
  ggplot() +
  geom_point(aes(baywide.count, section.count, color = alpha.code)) +
  facet_wrap(~section) +
  ggtitle("WEGR") +
  theme_bw()

```



For Widgeon, section 1 appears most important.

```{r error=FALSE, message=FALSE, echo=FALSE, warning=FALSE}

wbird_baywide_spp_season_survey %>% 
  filter(alpha.code %in% c("AMWI")) %>% 
  ggplot() +
  geom_point(aes(baywide.count, section.count, color = alpha.code)) +
  facet_wrap(~section) +
  ggtitle("AMWI") +
  theme_bw()

```


